---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  error = FALSE,
  message = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  dev = "svg"
)

ex1_alt <- paste("A bar chart with grey background, white grid lines",
                 "and dark grey bars.")
ex2_alt <- paste("A bar chart with white background,",
                 "light grey horizontal grid lines, and dark blue bars.")
ex3_alt <- paste("A line chart with white background,
                 light grey horizontal grid lines, an orange line, and a dark blue
                 line.")
```

# afcharts <img src="man/figures/logo.svg" alt="afcharts logo" align="right" height="150"/>

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/afcharts)](https://CRAN.R-project.org/package=afcharts)
[![R-CMD-check](https://github.com/best-practice-and-impact/afcharts/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/best-practice-and-impact/afcharts/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

afcharts is an R package for creating accessible plots by the Government Analysis Function. Currently, functions are available for styling ggplot2 plots.

The package has been developed using the [Government Analysis Function Data Visualisation guidance](https://analysisfunction.civilservice.gov.uk/policy-store/data-visualisation-charts/). afcharts should be used in conjunction with these guidance documents.

More information about the package and its functions can be found on the [afcharts website](https://best-practice-and-impact.github.io/afcharts/). In particular, the [cookbook](https://best-practice-and-impact.github.io/afcharts/articles/cookbook.html) contains lots of examples.

## Installation

### Install from CRAN

Install the latest release version of afcharts directly from CRAN:

``` {r install from CRAN, eval=FALSE}
install.packages("afcharts")
```


### Install from GitHub

afcharts can be installed directly from GitHub.

``` {r install from GitHub, eval=FALSE}
remotes::install_github(
  "best-practice-and-impact/afcharts",
  upgrade = "never",
  build_vignettes = TRUE,
  dependencies = TRUE
)
```

## Getting Started

Once installed, afcharts can be loaded using the `library()` function:

``` r
library(afcharts)
```

Help files for each function in the package can be found on the [References](https://best-practice-and-impact.github.io/afcharts/reference/) page of the package website. Alternatively, type `?function_name` into the RStudio console. For example:

``` r
?theme_af()
```

### Use afcharts as default

The easiest way to use afcharts is by adding `use_afcharts()` to the beginning of your R script, Rmarkdown document or Shiny app code. This function will set a number of defaults to ggplot2 geoms, use afcharts colour palettes and use `theme_af()`.

#### Example 1: Bar chart with one colour using ggplot2 defaults

```{r ex1, fig.alt = ex1_alt, fig.width = 10}
library(ggplot2)
library(dplyr)
library(gapminder)
library(afcharts)

gapminder |> 
  filter(year == 2007 & continent == "Americas") |>
  slice_max(order_by = pop, n = 5) |>
  ggplot() +
  geom_col(aes(x = reorder(country, -pop), y = pop)) +
  scale_y_continuous(
    labels = scales::label_number(scale = 1E-6),
    limits = c(0, 350E6),
    expand = expansion(mult = c(0, 0.1))
  ) +
  scale_fill_discrete_af("focus", reverse = TRUE) +
  labs(
    x = NULL,
    y = NULL,
    title = "The U.S.A. is the most populous country in\nthe Americas",
    subtitle = "Population of countries in the Americas (millions), 2007",
    caption = "Source: Gapminder"
    )
```

#### Example 2: Bar chart of one colour using afcharts defaults

```{r ex2, fig.alt = ex2_alt, fig.width = 10}
afcharts::use_afcharts()

gapminder |> 
  filter(year == 2007 & continent == "Americas") |>
  slice_max(order_by = pop, n = 5) |>
  ggplot(aes(x = reorder(country, -pop), y = pop)) +
  geom_col(fill = af_colour_values["dark-blue"]) +
  scale_y_continuous(
    labels = scales::label_number(scale = 1E-6),
    limits = c(0, 350E6),
    expand = c(0, 0),expansion(mult = c(0, 0.1))
  ) +
  labs(
    x = NULL,
    y = NULL,
    title = "The U.S.A. is the most populous country in\nthe Americas",
    subtitle = "Population of countries in the Americas (millions), 2007",
    caption = "Source: Gapminder"
  )
```

#### Example 3: Multiple colour line chart with afcharts formatting

```{r ex3, fig.alt = ex3_alt, fig.width = 10}
afcharts::use_afcharts()

gapminder |>
  filter(country %in% c("United Kingdom", "China")) |>
  ggplot(aes(x = year, y = lifeExp, colour = country)) +
  geom_line(linewidth = 1) +
  scale_y_continuous(
    breaks = seq(0, 80, 20),
    limits = c(0, 82),
    expand = expansion(mult = c(0, 0.1))
  ) +
  scale_x_continuous(breaks = seq(1952, 2007, 5)) +
  labs(
    x = "Year",
    y = NULL,
    title = "Living Longer",
    subtitle = "Life Expectancy in the United Kingdom and China 1952-2007",
    caption = "Source: Gapminder",
    colour = NULL
  )
```

**Note on use of titles, subtitles and captions** <br> Titles, subtitles and captions have been embedded in these example charts for demonstration purposes. However, for accessibility reasons, it is usually preferable to provide titles in the body of the page rather than embedded within the image of the plot. More information is available in the [accessibility article](https://best-practice-and-impact.github.io/afcharts/articles/accessibility.html#other-accessibility-considerations).

## Acknowledgments

The afcharts package is based on the [sgplot](https://scotgovanalysis.github.io/sgplot/index.html) package, written by Alice Hannah.

## Licence

Unless stated otherwise, the codebase is released under [the MIT License](LICENSE). This covers both the codebase and any sample code in the documentation.

The documentation is [© Crown copyright](https://www.nationalarchives.gov.uk/information-management/re-using-public-sector-information/uk-government-licensing-framework/crown-copyright/) and available under the terms of the [Open Government 3.0](https://www.nationalarchives.gov.uk/doc/open-government-licence/version/3/) licence.
