---
title: "Colours"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Colours}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  echo = FALSE
)

table_note <- paste(
  "Note: the cells in the ‘Example of colour’ column may appear blank",
  "to screen reader software. They contain a colour fill, but no data."
)
```

afcharts provides colour palettes as set out by the Government [Analysis Function Colours Guidance](https://analysisfunction.civilservice.gov.uk/policy-store/data-visualisation-colours-in-charts/). These palettes have been developed to meet the [Web Content Accessibility Guidelines 2.2 for graphical objects](https://www.w3.org/WAI/WCAG22/Understanding/non-text-contrast.html). 


## Things to consider when using colour

Using afcharts colour palettes does not guarantee that your charts will be accessible. 

The [Analysis Function Colours Guidance](https://analysisfunction.civilservice.gov.uk/policy-store/data-visualisation-colours-in-charts/) outlines other steps you should take to help make sure the colours you use in data visualisations are accessible.
afcharts should be used in conjunction with this advice.

The following are a few key things to consider when using colour in data visualisations:

*   Only use colour if absolutely necessary

*   Do not rely on colour alone to communicate a message

*   Use colour consistently when producing a series of plots

*   Limit the number of different colours you use; ideally an absolute maximum of four

*   Use colour palettes in the order presented in the guidance to ensure adjacent colours have sufficient contrast with each other


## Government Analysis Function colour palettes

To use an Analysis Function colours, set `palette_type = "af"` when using any of the afcharts colour or fill [`scale_` functions](https://best-practice-and-impact.github.io/afcharts/reference/index.html#colours). Set the `palette` argument to `"categorical"`, `"sequential"` or `"focus"` depending on which palette you wish to use.


### Categorical palette

The `categorical` palette should be used with categorical data. The colours in this palette all have at least a 3:1 colour contrast ratio with white (#ffffff), as well as with adjacent colours in the palette when used in the order recommended in the guidance. 

`r table_note`

```{r}
afcharts:::colour_table(afcharts::af_colour_palettes$categorical)
```

The afcharts categorical colour palette supports chart designs with up to 6 categorial values. Colours should be used in the order as shown above. The exception is when only two colours are required. In this scenario, dark blue and orange should be used instead, as these have the greatest colour contrast with each other. The afcharts discrete `scale_` functions will automatically use the correct colours based on the number you need.

`r table_note`

```{r}
afcharts:::colour_table(afcharts::af_colour_palettes$categorical2)
```


It is recommended that charts contain no more than 4 colours to represent categorical values.

### Sequential palette

The `sequential` colour palette should be used for data where the order has some meaning.

`r table_note`

```{r}
afcharts:::colour_table(afcharts::af_colour_palettes$sequential)
```

As this palette is restricted to tints of one colour, the colour contrast is not optimal.
Follow the [advice for using the sequential palette](https://analysisfunction.civilservice.gov.uk/policy-store/data-visualisation-colours-in-charts/#section-6) in the Government Analysis Function colour guidance when using this palette.

The afcharts discrete `scale_` functions will use the correct colours from this palette based on the number that you need.


### Focus palette

The `focus` palette should be used to highlight specific elements of a plot.

`r table_note`

```{r}
afcharts:::colour_table(afcharts::af_colour_palettes$focus)
```

The light grey colour in this palette does not have a strong colour contrast against a white background and therefore this palette should only be used when it is essential to communicate your message. 
Follow the [advice for using the focus palette](https://analysisfunction.civilservice.gov.uk/policy-store/data-visualisation-colours-in-charts/#section-7) in the Government Analysis Function colour Guidance when using this palette.


## Departmental colour palettes

afcharts does not support departmental branded colour palettes. The reasons for this decision are as follows:

- The afcharts team does not have the resources to individually QA colour palettes to ensure they meet the accessibility guidelines set out by the Government Analysis Function.
- It is difficult to create a colour palette that meets the accessibility guidelines and also looks substantially different to the existing palettes offered by afcharts.


## Using your own colour palette

There may be instances where you'd like to use a different colour palette. 
If so, this should be carefully considered to ensure it meets accessibility requirements. 
The Government Analysis Function guidance outlines [appropriate steps for choosing your own accessible colour palette](https://analysisfunction.civilservice.gov.uk/policy-store/data-visualisation-colours-in-charts/#section-9).

An example of how to use an alternative colour palette is provided in the [cookbook](https://best-practice-and-impact.github.io/afcharts/articles/cookbook.html#using-your-own-colour-palette).

If you use a different palette regularly and feel it would be useful for this to be added to afcharts, please make a suggestion as per the [contributing guidance](https://best-practice-and-impact.github.io/afcharts/CONTRIBUTING.html).


## Viewing palettes in R

Names and hex codes for available colours and palettes can be viewed by running the following code:

```{r view-colours, eval = FALSE, echo = TRUE}
# View names and hex codes for all colours
afcharts::af_colour_values

# View names and hex codes for all colour palettes
afcharts::af_colour_palettes

# View names and hex codes for `categorical` colour palette
afcharts::af_colour_palettes$categorical
```

Examples of how to apply these palettes to ggplot2 charts are available in both the [cookbook](https://best-practice-and-impact.github.io/afcharts/articles/cookbook.html) and the reference files for `scale_` [colour functions](https://best-practice-and-impact.github.io/afcharts/reference/index.html#colours).
