---
title: "Colours"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Colours}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  echo = FALSE
)

table_note <- paste(
  "Note: the cells in the ‘Example of colour’ column may appear blank",
  "to screen reader software. They contain a colour fill, but no data."
)
```

afcharts provides colour palettes as set out by the [Government Analysis Function](https://analysisfunction.civilservice.gov.uk/policy-store/data-visualisation-colours-in-charts/). 

These palettes have been developed to meet the [Web Content Accessibility Guidelines 2.2 for graphical objects](https://www.w3.org/WAI/WCAG22/Understanding/non-text-contrast.html). The colours in the [palette for categorical data](#categorical-palette) all have at least a 3:1 colour contrast ratio with white (#ffffff), as well as with adjacent colours in the palette when used in the order recommended in the guidance. 


## Things to consider when using colour

Using afcharts colour palettes does not guarantee that your charts will be accessible. 
The [Government Analysis Function](https://analysisfunction.civilservice.gov.uk) has published [guidance](https://analysisfunction.civilservice.gov.uk/policy-store/data-visualisation-colours-in-charts/) on other steps you should take to help make the colours you use in data visualisations more accessible.
afcharts should be used in conjunction with this advice.

The following are a few key things to consider when using colour in data visualisations:

*   Only use colour if absolutely necessary

*   Do not rely on colour alone to communicate a message

*   Use colour consistently when producing a series of plots

*   Limit the number of different colours you use; ideally an absolute maximum of four

*   Use colour palettes in the order presented to ensure adjacent colours have sufficient contrast with each other


## Government Analysis Function colour palettes

The Government Analysis Function guidance also contains [suggested colour palettes](https://analysisfunction.civilservice.gov.uk/policy-store/data-visualisation-colours-in-charts/#section-4). These are provided in afcharts. To use an Analysis Function palette, set `palette_type = "af"` when using any of the `scale_` [colour functions](https://best-practice-and-impact.github.io/afcharts/reference/index.html#colours).


### Categorical palette

The `categorical` palette should be used with categorical data.

`r table_note`

```{r}
afcharts:::colour_table(afcharts::af_colour_palettes$categorical)
```

The `categorical2` palette should be used if only two colours are required as the dark blue and orange colours have the greatest colour contrast with each other. 
If only two colours are required the `categorical` family is requested, `categorical2` will be used instead.

`r table_note`

```{r}
afcharts:::colour_table(afcharts::af_colour_palettes$categorical2)
```


### Sequential palette

The `sequential` colour palette should be used for data where the order has some meaning.

`r table_note`

```{r}
afcharts:::colour_table(afcharts::af_colour_palettes$sequential)
```

As this palette is restricted to tints of one colour, colour contrast is not optimal.
Follow the [advice for using the sequential palette](https://analysisfunction.civilservice.gov.uk/policy-store/data-visualisation-colours-in-charts/#section-6) in the Government Analysis Function colour guidance when using this palette.


### Focus palette

The `focus` palette should be used to highlight specific elements of a plot.

`r table_note`

```{r}
afcharts:::colour_table(afcharts::af_colour_palettes$focus)
```

The light grey colour in this palette does not have sufficient colour contrast against a white background and therefore this palette should only be used when it is essential to communicate your message. 
Follow the [advice for using the focus palette](https://analysisfunction.civilservice.gov.uk/policy-store/data-visualisation-colours-in-charts/#section-7) in the Government Analysis Function colour guidance when using this palette.


## Departmental colour palettes

afcharts does not support departmental colour palettes. The reasons for this decision are as follows:

- The afcharts team does not have the resources to individually QA colour palettes to ensure they meet the accessibility guidelines set out by the Government Analysis Function.
- It is difficult to create a colour palette that meets the accessibility guidelines and also looks substantially different to the existing palettes offered by afcharts.
- There are many different government departments, maintaining a custom palette for each is unrealistic.


## Using your own colour palette

There may be instances where you'd like to use a different colour palette. 
If so, this should be carefully considered to ensure it meets accessibility requirements. 
The Government Analysis Function guidance outlines [appropriate steps for choosing your own accessible colour palette](https://analysisfunction.civilservice.gov.uk/policy-store/data-visualisation-colours-in-charts/#section-9) and should be used.

An example of how to use an alternative colour palette is provided in the [cookbook](https://best-practice-and-impact.github.io/afcharts/articles/cookbook.html#using-your-own-colour-palette).
However, if you use a different palette regularly and feel it would be useful for this to be added to afcharts, please make a suggestion as per the [contributing guidance](https://best-practice-and-impact.github.io/afcharts/CONTRIBUTING.html).


## Viewing palettes in R

Names and hex codes for available colours and palettes can be viewed by running the following code:

```{r view-colours, eval = FALSE, echo = TRUE}
# View names and hex codes for all colours
afcharts::af_colour_values

# View names and hex codes for all colour palettes
afcharts::af_colour_palettes

# View names and hex codes for `categorical` colour palette
afcharts::af_colour_palettes$categorical
```

Examples of how to apply these palettes to ggplot2 charts are available in both the [cookbook](https://best-practice-and-impact.github.io/afcharts/articles/cookbook.html) and the reference files for `scale_` [colour functions](https://best-practice-and-impact.github.io/afcharts/reference/index.html#colours).


## Categorical palette ordering

The afcharts categorical colour palette supports chart designs with up to 6 categorial values. Colours are selected to represent categorical values in a pre-determined default order.

- For 2 categorical values, dark blue ("#12436D") is selected as the primary colour and orange ("#F46A25") as the secondary colour.
- For 3 categorical values, dark blue is selected as the primary colour, followed by turqouise ("#28A197") as the secondary colour and dark pink ("#801650") as the tertiary colour.
- For 4 categorical values, the colours are given in the order dark blue, turqouise, dark pink, and orange.
- For 5 categorical values, the colours are given in the order dark blue, turqouise, dark pink, orange, and dark grey (#3D3D3D").
- Finally, for 6 categorical values, the colours are given in the order dark blue, turqouise, dark pink, orange, dark grey, and light purple ("#A285D1").

It is recommened that charts contain no more than 4 different categorical values, and no more than 6 are supported by afcharts.

