## Using different colour palettes

afcharts provides colour palettes as set out by the [Government Analysis Function suggested colour palettes](https://analysisfunction.civilservice.gov.uk/policy-store/data-visualisation-colours-in-charts/#section-4). 
These palettes have been developed to meet the [Web Content Accessibility Guidelines 2.1 for graphical objects](https://www.w3.org/WAI/WCAG21/Understanding/non-text-contrast.html).


The [main palette](https://best-practice-and-impact.github.io/afcharts/articles/colours.html#main-palette) is the default for discrete colour/fill functions, and the [sequential palette](https://best-practice-and-impact.github.io/afcharts/articles/colours.html#sequential-palette) for continuous colour/fill functions.

More information on the colours used in afcharts can be found at `vignette("colours")`.

### Using afcharts colour palettes

The full list of available palettes can be found by running `afcharts::af_colour_palettes`.

For example, to use the Analysis Function `main2` palette:

```{r af-palette, eval = FALSE}
gapminder |>
  filter(country %in% c("United Kingdom", "China")) |>
  ggplot(aes(x = year, y = lifeExp, colour = country)) +
  geom_line(linewidth = 1) +
  theme_af(legend = "bottom") +
  scale_colour_discrete_af("main2") +
  scale_y_continuous(
    limits = c(0, 82),
    breaks = seq(0, 80, 20),
    expand = c(0, 0)
  ) +
  scale_x_continuous(breaks = seq(1952, 2007, 5)) +
  labs(
    x = "Year",
    y = NULL,
    colour = NULL
  )
```

<div class="chart-title">Living Longer</div>
<div class="chart-subtitle">Life Expectancy in the United Kingdom and China 1952-2007</div>
```{r, fig.height = 5, echo = FALSE}
<<af-palette>>
```

This line chart uses the afcharts theme and the Analysis Function main palette, with one line in dark blue and one in orange.

### Using your own colour palette

There may be instances where you'd like to use a colour palette that is not available in afcharts. 
If so, this should be carefully considered to ensure it meets accessibility requirements. 
The Government Analysis Function guidance outlines [appropriate steps for choosing your own accessible colour palette](https://analysisfunction.civilservice.gov.uk/policy-store/data-visualisation-colours-in-charts/#section-9) and should be used.

```{r different-colour-palette-1, eval = FALSE}
my_palette <- c("#0F820D", "#000000")

gapminder |>
  filter(country == "United Kingdom") |>
  ggplot(aes(x = year, y = lifeExp)) +
  geom_line(linewidth = 1, colour = my_palette[1]) +
  theme_af() +
  scale_y_continuous(
    limits = c(0, 82),
    breaks = seq(0, 80, 20),
    expand = c(0, 0)
  ) +
  scale_x_continuous(breaks = seq(1952, 2007, 5)) +
  labs(
    x = "Year",
    y = NULL,
  )
```

<div class="chart-title">Living Longer</div>
<div class="chart-subtitle">Life Expectancy in the United Kingdom 1952-2007</div>
```{r, echo = FALSE}
<<different-colour-palette-1>>
```

This line chart uses the afcharts theme with a single green line.

Alternatively, you may wish to use multiple custom colours:

```{r different-colour-palette-2, eval = FALSE}
gapminder |>
  filter(country %in% c("United Kingdom", "China")) |>
  ggplot(aes(x = year, y = lifeExp, colour = country)) +
  geom_line(linewidth = 1) +
  theme_af(legend = "bottom") +
  scale_colour_manual(values = my_palette) +
  scale_y_continuous(
    limits = c(0, 82),
    breaks = seq(0, 80, 20),
    expand = c(0, 0)
  ) +
  scale_x_continuous(breaks = seq(1952, 2007, 5)) +
  labs(
    x = "Year",
    y = NULL,
    colour = NULL
  )
```

<div class="chart-title">Living Longer</div>
<div class="chart-subtitle">Life Expectancy in the United Kingdom and China 1952-2007</div>
```{r, fig.height = 5.5, echo = FALSE}
<<different-colour-palette-2>>
```

This line chart uses the afcharts theme. One line is green and the other is black.


#### Adding a new colour palette to afcharts

If you use a different palette regularly and feel it would be useful for this to be added to afcharts, please make a suggestion as per the [contributing guidance](https://best-practice-and-impact.github.io/afcharts/CONTRIBUTING.html).
